\Large Inferential Analysis of the Supreme Court Citation Network

### 2018-04-09 Tex/paper.tex

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part to opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citaitons are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
Prepared for presentation at the ``The Enemy of My Enemy: Judicial Politics meets Network Analysis'' conference within a conference at the 2018 Annual Conference of the Southern Political Science Association. This work was supported in part by NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219.}
\end{abstract}

@@

\noindent \textbf{Popularity} Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed.  which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. 
\begin{quotation}
The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 times the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

@@

 \section{The Citation Temporal Exponential Random Graph Model}
We build upon existing methods for network analysis to define a model that fits the constraints to which Supreme Court citation networks are subject. The Supreme Court citation network is nearly a directed acyclic graph---a graph in which there can be no loops/cycles (e.g., ties cannot be reciprocated, which would constitute a two-node loop). If two cases are decided in different terms, the later case can cite the earlier case, but not the reverse. However, as noted above, if two opinions are written in the same term, they can cite each other reciprocally. We implement a version of the temporal exponential random graph model (TERGM)---a model that is commonly used for longitudinal network data \citep[e.g.,][]{cranmer2012complex,desmarais2013forecasting,clark2013multimember,block2018change,graif2017neighborhood}---that is subject to the constraints specific to the Supreme Court citation network. The TERGM is a model that can simultaneously incorporate structural dependencies such as transitivity and reciprocity and exogenous covariates  (e.g. the issue area of the citing case, the ideological distance between the opinion authors of the two cases in the dyad) to explain tie formation. If the ties do not depend upon each other, and the coefficients associated with the structural dependencies are either assumed or estimated to be zero, the TERGM reduces to a logistic regression in which the observation is a directed dyad \citep{cranmer2010inferential}.  In this section we describe the citation TERGM (c-TERGM), and explain estimation methods.

Let $C_t$ be the case-to-case adjacency matrix at time $t$ such that $C_t^{ij} \in \{0,1\}$ is a binary indicator of whether case $i$ cites $j$. Furthermore, let
$$\mathcal{C}_t =\{C_{\leq t} \in \{0,1\}^{(N_t \cdot (N_t-1))/2}: C_{\leq t}^{ij} \in \{0,1\} \}$$ 
be the set of all possible adjacency matrices among $N_t$ cases at time $t$. $N_t$ denotes the the number of cases in the network at time $t$. Note that the cardinality of $\mathcal{C}_t$ increases exponentially for every newly added case. The probability of observing $C_t$ given past citations $C_{<t}$, where $C_{\leq t}$ is the network up to time $t$ is defined as
\begin{equation}
P(C_t | C_{<t}, \theta)=\cfrac{\exp\big(\theta^T\cdot h\big(C_{\leq t})\big)\big)}{\sum_{Z_t^*\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)\big)}
\label{ercm}
\end{equation}
where $\theta \in \mathbb{R}^q$ is a $q$-dimensional vector of parameters,  $h: \mathcal{C}_t \to \mathbb{R}^q$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{Z^*_t\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)$ is a normalization constant that ensures that Equation (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.

@@

We include two sets of dummy variables that account for the issue areas of cases. In one set of dummy variable---sender intercepts---the variable $X_{ij}$ indicates the issue area of the sending case ($i$). In the set of receiver intercepts $X_{ij}$ indicates the issue are of the receiver case. Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} find that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} states which number represents which issie area.

@@

 \section{Empirical Analysis}
Our three data sources for this study include the Supreme Court Database (SCDB) \citep{spaeth2014supreme}, Martin-Quinn scores \citep{martin2002dynamic}, and Supreme Court citation data provided by the CourtListener Free Law Project \citep{CourtListener}. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015)\footnote{There were 145 cases that were listed in the SCDB but could not be matched to a case in the Courtlistener data. We decided to exclude these 145 cases from our model.}. The Supreme Court citation network from $1937 - 2015$ consists of $9,945$ cases. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. The network has a total of $111,986$ ties. The number of triangles in the network is $250,717$ and the number of mutual ties is $56$. The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in Figure \ref{degree_dist}. The maximum indegree is $230$ and the maximum outdegree is $162$.

@@

\begin{itemize}
\item update Figure 2 to time period 1937 - 2015
\end{itemize}

@@

\begin{table}[H]
\centering
\begin{tabular}{rllll}
\hline \hline
& \multicolumn{2}{l}{Independent Model } & \multicolumn{2}{l}{Full Model} \\ \hline 
 & mean & range & mean & range \\ 
  \hline
precision & 0.5872 & (0.5781, 0.5962) & 0.8622 & (0.8593, 0.8668) \\ 
  recall & 0.0565 & (0.0547, 0.0587) & 0.5839 & (0.5777, 0.592) \\ 
  F1 score & 0.103 & (0.1, 0.1069) & 0.6963 & (0.6915, 0.7027) \\ 
   \hline \hline
\end{tabular}
\caption{The predictive performance of the directed dyadic models with no trends, over ten 80/20 train/test splits.}
\label{tab:prediction_notrend}
\vspace{0.4cm}
\begin{tabular}{rllll}
\hline \hline
& \multicolumn{2}{l}{Independent Model } & \multicolumn{2}{l}{Full Model} \\ \hline 
 & mean & range & mean & range \\ 
  \hline
precision & 0.6441 & (0.6302, 0.6536) & 0.8671 & (0.8626, 0.873) \\ 
  recall & 0.0852 & (0.0832, 0.0863) & 0.594 & (0.5874, 0.6012) \\ 
  F1 score & 0.1504 & (0.1473, 0.1525) & 0.705 & (0.7008, 0.712) \\ 
   \hline \hline
\end{tabular}
\caption{The predictive performance of the directed dyadic models with quadratic trends, over ten 80/20 train/test splits.}
\label{tab:prediction_quadratic}
\end{table}
We also conduct a predective experiment for the model that does not consider any trends and for the model that does consider quadratic instead of linear trends. The predictive performance results are given in tables \ref{tab:prediction_notrend} and \ref{tab:prediction_quadratic}. By comparing the full model results in tables \ref{tab:prediction_notrend} and \ref{tab:prediction_quadratic} with the results of the full linear trend model in table \ref{tab:prediction_linear} we can conclude all three models perform similarly well.   \\[0,5cm]

>>>>

\maketitle
\begin{abstract}
\noindent The significance and influence of US Supreme Court majority opinions derive in large part to opinions' roles as precedents for future opinions. A growing body of literature seeks to understand what drives the use of opinions as precedents through the study of Supreme Court case citation patterns. We raise two limitations of existing work on Supreme Court citations. First, dyadic citations are typically aggregated to the case level before they are analyzed. Second, citations are treated as if they arise independent of one another. We make the case that studying Supreme Court citations as a dynamic citation network overcomes both of these limitations. We present a methodology for studying citations between US Supreme Court opinions at the dyadic level, as a network. This methodology---the citation temporal exponential random graph model---enables researchers to account for the effects of case characteristics and complex forms of network dependence in citation formation. We apply this methodology to a network that includes all Supreme Court cases decided between 1937 and 2015. We find evidence for dependence processes, including reciprocity, transitivity, popularity, and activity. The dependence effects that we identify are as substantively and statistically significant as the effects of the exogenous covariates we include in the model. The summary result from this analysis is that theoretical models of Supreme Court citation formation should consider both the effects of case characteristics and the structure of past citations.\footnote{
Prepared for presentation at the ``The Enemy of My Enemy: Judicial Politics meets Network Analysis'' conference within a conference at the 2018 Annual Conference of the Southern Political Science Association. This work was supported in part by NSF grants SES- 1558661, SES-1637089, SES-1619644, and CISE-1320219.}
\end{abstract}

@@

\noindent \textbf{Popularity} Popularity, also termed ``preferential attachment'' is the tendency for ties to be sent to nodes to which many ties have already been sent \citep{barabasi1999emergence}. Citations to an opinion signal both the Court's awareness of the legal reasoning of the case and the Court's evaluation that the opinion is an authoritative precedent. The more citations, the stronger this signal. Landmark cases, or those that establish new legal rules, are particularly authoritative and accrue citations from most future opinions that follow the respective line of reasoning. The passage below, from Oregon v. Mitchell (400 U.S. 112, 1970)---a case on the legality of state age restrictions on voting in federal elections---illustrates this popularity dynamic. In this opinion passage Baker v. Carr is cited in reference to its role as a landmark precedent, and noted for the number of other cases by which it has been followed and for which an authoritative opinion is referenced, and even discussed in terms of the number of other cases by which it was followed. 
\begin{quotation}
The first case in which this Court struck down a statute under the Equal Protection Clause of the Fourteenth Amendment was Strauder v. West Virginia, 100 U. S. 303, decided in the 1879 Term. [Footnote 2/1] In the 1961 Term, we squarely held that the manner of apportionment of members of a state legislature raised a justiciable question under the Equal Protection Clause, Baker v. Carr, 369 U. S. 186. That case was followed by numerous others, e.g.: that one person could not be given twice or 10 times the voting power of another person in a state-wide election merely because he lived in a rural area..."
\end{quotation} %https://supreme.justia.com/cases/federal/us/400/112/case.html

@@

 \section{The Citation Temporal Exponential Random Graph Model}
We build upon existing methods for network analysis to define a model that fits the constraints to which Supreme Court citation networks are subject. The Supreme Court citation network is nearly a directed acyclic graph---a graph in which there can be no loops/cycles (e.g., ties cannot be reciprocated, which would constitute a two-node loop). If two cases are decided in different terms, the latter case can cite the earlier case, but not the reverse. However, as noted above, if two opinions are written in the same term, they can cite each other reciprocally. We implement a version of the temporal exponential random graph model (TERGM)---a model that is commonly used for longitudinal network data \citep[e.g.,][]{cranmer2012complex,desmarais2013forecasting,clark2013multimember,block2018change,graif2017neighborhood}---that is subject to the constraints specific to the Supreme Court citation network. The TERGM is a model that can simultaneously incorporate structural dependencies such as transitivity and reciprocity and exogenous covariates  (e.g. the issue area of the citing case, the ideological distance between the opinion authors of the two cases in the dyad) to explain tie formation. If the ties do not depend upon each other, and the coefficients associated with the structural dependencies are either assumed or estimated to be zero, the TERGM reduces to a logistic regression in which the observation is a directed dyad \citep{cranmer2010inferential}.  In this section we describe the citation TERGM (c-TERGM), and explain estimation methods.

Let $C_t$ be the case-to-case adjacency matrix at time $t$ such that $C_t^{ij} \in \{0,1\}$ is a binary indicator of whether case $i$ cites $j$. Furthermore, let
$$\mathcal{C}_t =\{C_{\leq t} \in \{0,1\}^{(N_t \cdot (N_t-1))/2}: C_{\leq t}^{ij} \in \{0,1\} \}$$ 
be the set of all possible adjacency matrices among $N_t$ cases at time $t$. $N_t$ denotes the number of cases in the network at time $t$. Note that the cardinality of $\mathcal{C}_t$ increases exponentially for every newly added case. The probability of observing $C_t$ given past citations $C_{<t}$, where $C_{\leq t}$ is the network up to time $t$ is defined as
\begin{equation}
P(C_t | C_{<t}, \theta)=\cfrac{\exp\big(\theta^T\cdot h\big(C_{\leq t})\big)\big)}{\sum_{Z_t^*\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)\big)}
\label{ercm}
\end{equation}
where $\theta \in \mathbb{R}^q$ is a $q$-dimensional vector of parameters,  $h: \mathcal{C}_t \to \mathbb{R}^q$ is a q-dimensional vector of different statistics and $\kappa(\theta) := \sum_{Z^*_t\in \mathcal{C}_t}\exp\big(\theta^T\cdot h\big(Z^*_{\leq t})\big)$ is a normalization constant that ensures that Equation (\ref{ercm}) defines a probability function on $\mathcal{C}_t$.

@@

We include two sets of dummy variables that account for the issue areas of cases. In one set of dummy variable---sender intercepts---the variable $X_{ij}$ indicates the issue area of the sending case ($i$). In the set of receiver intercepts $X_{ij}$ indicates the issue are of the receiver case. Issue area data comes from the Supreme Court Database (SCDB) \citep{spaeth2014supreme}. We include these variables because \citet{cross2010determinants} find that the number of citations to Supreme Court opinions depends heavily on the issue area of the case. Table \ref{issue_area_coding} states which number represents which issue area.

@@

 \section{Empirical Analysis}
Our three data sources for this study include the Supreme Court Database (SCDB) \citep{spaeth2014supreme}, Martin-Quinn scores \citep{martin2002dynamic}, and Supreme Court citation data provided by the CourtListener Free Law Project \citep{CourtListener}. We limit the Supreme Court terms included in our analysis to those that are covered by all three of these data source (1937--2015)\footnote{There were 145 cases that were listed in the SCDB but could not be matched to a case in the CourtListener data. We decided to exclude these 145 cases from our model.}. The Supreme Court citation network from $1937 - 2015$ consists of $9,945$ cases. The breakdown of the data by the Court's Chief Justice is presented in Table \ref{tab:chiefs}. In order to only account for positive citations, we exclude $315$ citations that caused the cited case to be overruled. As a result, the network has a total of $111,986$ ties. The number of triangles in the network is $250,717$ and the number of mutual ties is $56$. The in- and outdegree distributions (i.e., the distributions of the number of citations sent and received by cases, respectively), are visualized in Figure \ref{degree_dist}. The maximum indegree is $230$ and the maximum outdegree is $162$.

@@

\begin{itemize}
\item update Figure 2 to time period 1937 - 2015
\item add data source for overruling citations
\end{itemize}

@@

\begin{table}[H]
\centering
\begin{tabular}{rllll}
\hline \hline
& \multicolumn{2}{l}{Independent Model } & \multicolumn{2}{l}{Full Model} \\ \hline 
 & mean & range & mean & range \\ 
  \hline
precision & 0.5872 & (0.5781, 0.5962) & 0.8622 & (0.8593, 0.8668) \\ 
  recall & 0.0565 & (0.0547, 0.0587) & 0.5839 & (0.5777, 0.592) \\ 
  F1 score & 0.103 & (0.1, 0.1069) & 0.6963 & (0.6915, 0.7027) \\ 
   \hline \hline
\end{tabular}
\caption{The predictive performance of the directed dyadic models with no trends, over ten 80/20 train/test splits.}
\label{tab:prediction_notrend}
\vspace{0.4cm}
\begin{tabular}{rllll}
\hline \hline
& \multicolumn{2}{l}{Independent Model } & \multicolumn{2}{l}{Full Model} \\ \hline 
 & mean & range & mean & range \\ 
  \hline
precision & 0.6441 & (0.6302, 0.6536) & 0.8671 & (0.8626, 0.873) \\ 
  recall & 0.0852 & (0.0832, 0.0863) & 0.594 & (0.5874, 0.6012) \\ 
  F1 score & 0.1504 & (0.1473, 0.1525) & 0.705 & (0.7008, 0.712) \\ 
   \hline \hline
\end{tabular}
\caption{The predictive performance of the directed dyadic models with quadratic trends, over ten 80/20 train/test splits.}
\label{tab:prediction_quadratic}
\end{table}
We also conduct a predictive experiment for the model that does not consider any trends and for the model that does consider quadratic instead of linear trends. The predictive performance results are given in tables \ref{tab:prediction_notrend} and \ref{tab:prediction_quadratic}. By comparing the full model results in tables \ref{tab:prediction_notrend} and \ref{tab:prediction_quadratic} with the results of the full linear trend model in table \ref{tab:prediction_linear} we can conclude all three models perform similarly well.   \\[0,5cm]
